Rotating Stars from \Kepler Observed with Gaia DR2

### 2018-07-20 paper/ms.tex
comments round 2 from KRC

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler. Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone, removing many sub-giants and unresolved binary stars from the sample.
The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is clearly recovered for stars out to 525pc, but is not detectable at further distances. This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence. The diagonal offset from the main sequence is unexpected, however, and may provide a new independent test of both stellar evolution and gyrochronlogy models.
\end{abstract}

@@

%\clearpage
%%%%%%%%%%%%%%%%%%%%%%
\section{Recalibrating Stellar Evolution Models}

\begin{figure*}
\centering
\includegraphics[width=6in]{../figures/cmd_zoom}
\caption{
Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with main sequence stars colored by their measured rotation periods from \citet{mcquillan2014}.
MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines). The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles). 
Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel). Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period. Error bars shown are the standard deviation of $\Delta M_G$ in each bin. The 0.7 M$_\odot$ star brightness evolution is shown as a function of rotation evolution from \citet{meibom2009} (black triangles).
}
\label{fig:cmd_zoom}
\end{figure*}

A subtle feature we noticed in Figure \ref{fig:cmd} is the color gradient (i.e. rotation period gradient) for the stars in between the single and binary star main sequence populations. In Figure \ref{fig:cmd} this appears as a yellow stripe (i.e. rotation periods of 30-40 days) between these blue-green sequences for systems with colors of $G_{BP} - G_{RP} \approx 1.5$. To exaggerate this feature, we have reproduced a portion of our color--magnitude diagram focused on the main sequence near this stellar color in the left panel of Figure \ref{fig:cmd_zoom}. A clear color gradient is present, with red points (slower rotators) appearing preferentially above the main sequence.
The right panel of Figure \ref{fig:cmd_zoom} concretely demonstrates a correlation between the measured rotation period and the vertical offset (i.e. absolute magnitude) from the $10^9$ year MIST isochrone. Slower rotating stars are brighter at a given color.

We interpret this observation as being due to the increase in luminosity of stars across their main sequence lives, coupled with the angular momentum loss that underpins gyrochronology. Indeed, the three MIST isochrones shown in Figure \ref{fig:cmd_zoom} with ages of $10^8$, $10^9$, and $10^{10}$ years clearly predict a net increase in brightness of the main sequence over time. We also track the color--magnitude evolution of a 0.7 $M_\odot$ star over its lifetime in Figure \ref{fig:cmd_zoom}, which shows the ``hook'' the star takes over its main sequence life.

Inverting the color--magnitude diagram into precise ages for main sequence stars using {\em Gaia} is beyond scope of this work. However, Figure \ref{fig:cmd_zoom} clearly demonstrates that age information is not only present for main sequence stars, but it may also be a critical factor in accurately modeling the CMD. For example: younger, presumably more metal rich stars, should have redder main sequence colors. This in turn should move the entire main sequence track to the right in the CMD. However, we find older stars (slower rotators) appearing on the right side of the main sequence, due to their increased luminosity with age. The main sequence location of a star is therefore not simply due to its mass and metallicity in the {\em Gaia} era, but its age must also be taken into account.

In Figure \ref{fig:cmd_zoom} we also show each stars' offset from the main sequence as a function of its rotation period within our zoomed-in CMD, as well as a prediction for the evolution of a 0.7 $M_\odot$ star. The spin-down model here is a combination of the MIST isochrone evolution, as well as the \citet{meibom2009} gyrochronology model. Note this spin-down model has been arbitrarily offset in $\Delta M_G$ to approximately match the observed rapid rotators. 
While \Kepler light curves are not able to measure rotation reliably for stars with periods longer than $\sim$30 days, we find that stars with brightness (vertical) offsets in the CMD consistent with being several Gyr old have average rotation periods far shorter than the 60--80 days values predicted from the spin-down models.

One interpretation of this result may be that the older stars are spinning {\it faster} than expected for stars their age. This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}. Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long rotation periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.

In tracking only the brightness evolution of stars as they age, we neglect the color evolution shown in Figure \ref{fig:cmd_zoom} (left panel), which predicts a star will move roughly {\it along} the main sequence track as it changes in luminosity over time. Instead we clearly observe a diagonal gradient in the rotation periods consistent with being a stellar age affect, which may point to missing physics in the MIST models. 
We believe simultaneous constraints on both the precise CMD location and surface rotation rate for stars, as demonstrated here, will be a valuable utility for calibrating a new generation of isochrone and gyrochronology models.

%\clearpage
%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Using a sample of 16,248 single main sequence stars with measured rotation periods from \Kepler and parallaxes from {\em Gaia} DR2, we have begun to explore the spatial distribution of stellar ages near the Sun using gyrochronology. The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low galactic heights, rather than in a bubble centered around the Sun. This is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the galactic disk.

Unfortunately the \Kepler survey only covers a single pointing, and so our volume analysis of the stellar ages is limited here. However, with more than 15 distinct lines of sight available from K2, the extended \Kepler mission, we will soon be able to more than double our sample of rotating field stars and probe a much wider range of stellar ages. This may enable us to determine the spatial scales that star formation histories are coherant over in the nearby Milky Way, and make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}. For example, gyrochronology ages for disk stars within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g. see][]{choi2015}.

@@

Finally, in the {\em Gaia} era, stellar evolution modelers have the daunting task to explain an increasing number of precisely determined features from survey observations, such as the fascinating new main sequence gap revealed by \citet{jao2018}, and the main sequence width seen here in Figure \ref{fig:cmd}.
This work has presented strong motivation for a new generation of stellar evolution and rotation models to be produced to accurately model the {\em Gaia} main sequence. With \Kepler, K2, and soon TESS, rotation measurements will be available for hundreds of thousands of nearby stars that {\em Gaia} has provided precise distances for. We hope new isochrone models will reproduce both the observed colors and luminosities of stars due to their ages and abundances, as well as the evolution of their surface rotations.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler. Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars centered within a 0.5 mag region about a 1 Gyr isochrone, removing many sub-giants and unresolved binary stars from the sample.
The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is clearly recovered for stars out to 525pc, but is not detectable at further distances. This bimodality correlates with Galactic height as well, dropping strongly for stars above $Z>90$ pc.
We also find a significant width in the stellar main sequence of $\Delta M_G\sim$0.25 mag, as well as a coherent gradient of increasing rotation periods diagonal to the main sequence.
We interpret this to be the simultaneous change in luminosity and loss of angular momentum as stars evolve along the main sequence. The diagonal offset from the main sequence is unexpected, however, and may provide a new independent test of both stellar evolution and gyrochronlogy models.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Recalibrating Stellar Evolution Models}

\begin{figure*}
\centering
\includegraphics[width=7in]{../figures/cmd_zoom}
\caption{
{\bf Left:} Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with main sequence stars colored by their measured rotation periods from \citet{mcquillan2014}.
MIST isochrones at ages of $10^8$, $10^9$, and $10^{10}$ yr are shown for comparison (blue, black, and red lines). The predicted evolution of a 0.7 M$_\odot$ star is highlighted (black triangles). 
{\bf Center:} Difference in $M_G$ from the $10^9$ yr MIST isochrone as a function of rotation period (point color again indicates rotation period, as in Left panel). An increase in the median brightness is seen in bins of rotation period (blue squares). Error bars are the standard deviation of $\Delta M_G$ in each bin. The 0.7 M$_\odot$ star brightness evolution from MIST is shown as a function of rotation evolution from \citet{meibom2009}, manually shifted to match the rapid rotating stars (black triangles).
{\bf Right:} Same as the center panel, but representing the change in $G_{BP}-G_{RP}$ color as a function of measured rotation period. Again the MIST--Meibom model evolution (black triangles) has been shifted to match the rapid rotators. Stars are predicted to get {\it bluer} at older ages, rather than redder as our sample indicates.
}
\label{fig:cmd_zoom}
\end{figure*}

A subtle feature we noticed in Figure \ref{fig:cmd} is the {\it diagonal} color gradient (i.e. rotation period gradient) for the stars in between the single and binary star main sequence populations. In Figure \ref{fig:cmd} this appears as a yellow stripe (i.e. rotation periods of 30-40 days) between these blue-green sequences for systems with colors of $G_{BP} - G_{RP} \approx 1.5$. To exaggerate this feature, we have reproduced a portion of our color--magnitude diagram focused on the main sequence near this stellar color in the left panel of Figure \ref{fig:cmd_zoom}. A clear color gradient is present, with red points (slower rotators) appearing preferentially above and to the right of the main sequence.
The center panel of Figure \ref{fig:cmd_zoom} demonstrates a correlation between the measured rotation period and the vertical offset (i.e. absolute magnitude) from the $10^9$ year MIST isochrone. Slower rotating stars are brighter at a given color.
However, as the right panel in Figure \ref{fig:cmd_zoom} shows, the MIST models predict a bluer color for stars at older ages, while our data show the slower rotators have redder colors on average.

In the center and right panels of Figure \ref{fig:cmd_zoom} we show a prediction of the brightness and color evolution of a 0.7 $M_\odot$ star as a function of its rotation period over time. The predicted evolution here is a combination of the MIST isochrone models from $10^8$ to $10^{10}$ years, as well as the \citet{meibom2009} gyrochronology model over this same time window. Note this  combined model has been arbitrarily offset in $\Delta M_G$ and $\Delta(G_{BP} - G_{RP})$ in the middle and right panels, respectively, to approximately match the CMD position observed for the rapid rotators.

These data challenge our main sequence model evolution in two ways. Firstly, the diagonal gradient from the main sequence in rotation period is in conflict with the evolution predicted from MIST. The stellar evolution model shows that a star should evolve essentially {\it along} the main sequence track, as in the left panel of Figure \ref{fig:cmd_zoom}. The tension with our observed sample is highlighted in the right panel of Figure \ref{fig:cmd_zoom}, where the predicted colors for old stars get bluer, while our slice through the observed rotation period sample shows older stars being redder.

This tension in the CMD evolution is compounded when we consider stellar metallicity, which we have not varied in our MIST model realization. Younger, presumably more metal rich stars should have redder optical colors on the main sequence, while older metal poor stars should be more blue. Instead we find that older, slower rotating stars are redder and brighter than their rapidly rotating counterparts. While we do not have independent metallicity constraints for the \Kepler stars in this sample, Figure \ref{fig:cmd_zoom} shows that age may have as big of a factor as metallicity in the precise CMD location of main sequence stars. Likewise, stellar evolution models do not appear to reproduce the CMD positions for main sequence, low-mass stars of all ages.

The second, more subtle challenge presented by these data is the rotation periods observed for presumably older stars. 
We find that stars having $M_G$ offsets in the center panel of Figure \ref{fig:cmd_zoom} consistent with being several Gyr old have average rotation periods of 30--40 days, far shorter than the 60+ days values predicted by the spin-down model for a 0.7 $M_\odot$ star.  Note, \Kepler light curves are often not able to reliably measure rotation periods longer than $\sim$30 days. This bias means we may be missing even slower rotators from our \Kepler--{\em Gaia} combined sample, and do not know what the $M_G$ offset for such stars would be.

One possible interpretation of this result as observed here is that the older stars are spinning {\it faster} than expected for stars their age. This is qualitatively similar to the model of broken spin-down occurring at a critical Rossby number suggested by \citet{van-saders2016}.Though we cannot definitively confirm such intriguing rotation evolution from this initial investigation given the observation bias for long rotation periods from \Kepler, matching the {\em Gaia} CMD with other rotation period measurements may provide an ideal dataset to test the \citet{van-saders2016} model against.

%%%%%%%%%%%%%%%%%%%%%%
\vspace{0.5in}
\section{Discussion}

Using a sample of 16,248 single main sequence stars with measured rotation periods from \Kepler and parallaxes from {\em Gaia} DR2, we have begun to explore the spatial distribution of stellar ages near the Sun using gyrochronology. The bimodality in rotation periods first reported by \citet{mcquillan2013} appear to be constrained to low Galactic scaleheights, rather than in an obvious bubble centered around the Sun. Since height above the Galactic plane is assumed to be related to age, this is consistent with the rotation period bimodality being a direct tracer of the star formation history, and indicates a burst of star formation within the past $\sim$600 Myr within the Galactic disk.

Unfortunately the \Kepler survey only covers a single pointing, and so our volume analysis of the stellar ages is limited here. However, with 16 distinct lines of sight available from K2, the extended \Kepler mission, we will be able to more than double our sample of rotating field stars and probe a much wider range of stellar ages. This may enable us to determine the spatial scales over which star formation histories are coherant in the Milky Way disk. We could then make comparisons to the $\sim$100 pc resolved star formation history maps available for nearby galaxies such as Andromeda \citep[][]{lewis2015,williams2017}. For example, gyrochronology ages for disk stars within 1 kpc of the Sun could be used to test if star formation propagates due to spiral arm density waves \citep[e.g. see][]{choi2015}.

@@

Finally, in the {\em Gaia} era, stellar evolution modelers have the daunting task of explaining an increasing number of precisely determined features in the CMD.
This includes the fascinating new main sequence gap revealed by \citet{jao2018}, as well as the diagonal age (or rotation) gradient in the main sequence shown in our Figures \ref{fig:cmd} and \ref{fig:cmd_zoom}.
Our work presents strong motivation for a new generation of stellar evolution and rotation models to accurately reproduce the main sequence as observed with \Kepler and {\em Gaia}. With \Kepler, K2, and soon TESS, precise rotation measurements will be available for hundreds of thousands of nearby stars that {\em Gaia} has provided reliable distances for. We hope these data will guide new isochrone models in reproducing both the observed colors and luminosities of stars due to their ages and abundances, as well as the evolution of their surface rotation rates.
